---
title: "Hands-on Exercise 6: Geographical Segmentation with Spatially Constrained Clustering Techniques"
author: "Kang-Chih Huang"
date: "2026-10-01"
date-modified: "last-modified"
format: html
execute:
  warning: false
  freeze: auto
---

## Getting started

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, corrplot, ggpubr, heatmaply, GGally, cluster, ClustGeo, factoextra, NbClust, psych)
```

## Importing geospatial data

```{r}
shan_sf <- st_read(dsn = "data/geospatial",
                    layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)"))
```

## Importing aspatial data

```{r}
ict <- read_csv("data/aspatial/Shan-ICT.csv")
```

## Deriving ICT penetration rate

```{r}
ict_derived <- ict %>%
  mutate(RADIO_PR = Radio / `Total households` * 1000,
         TV_PR = Television / `Total households` * 1000,
         LLPHONE_PR = `Land line phone` / `Total households` * 1000,
         MPHONE_PR = `Mobile phone` / `Total households` * 1000,
         COMPUTER_PR = Computer / `Total households` * 1000,
         INTERNET_PR = `Internet at home` / `Total households` * 1000) %>%
  rename(TS_PCODE = `Township Pcode`)
```

## Joining spatial and aspatial data

```{r}
shan_sf <- left_join(shan_sf, ict_derived, by = "TS_PCODE")
```

## Exploratory data analysis

```{r}
ggplot(data = ict_derived, aes(x = RADIO_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "lightblue") +
  labs(title = "Distribution of Radio penetration rate")
```

```{r}
ggplot(data = ict_derived, aes(x = RADIO_PR)) +
  geom_boxplot(color = "black", fill = "lightblue")
```

```{r fig.width=10}
tm_shape(shan_sf) +
  tm_polygons(fill = "RADIO_PR",
              fill.scale = tm_scale_intervals(
                style = "jenks",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(title = "Radio penetration rate")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Radio penetration rate by township")
```

## Correlation analysis

```{r}
cluster_vars.cor <- cor(ict_derived[, 12:17])
corrplot.mixed(cluster_vars.cor,
                lower = "ellipse",
                upper = "number",
                tl.pos = "lt",
                diag = "l",
                tl.col = "black")
```

## Hierarchical clustering

### Extracting clustering variables

```{r}
cluster_vars <- shan_sf %>%
  st_drop_geometry() %>%
  select(`Township Name`, RADIO_PR, TV_PR, LLPHONE_PR, MPHONE_PR, COMPUTER_PR) %>%
  rename(TS_NAME = `Township Name`)

row.names(cluster_vars) <- cluster_vars$TS_NAME
shan_ict <- select(cluster_vars, -TS_NAME)
```

### Standardising the clustering variables

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

### Computing proximity matrix

```{r}
proxmat <- dist(shan_ict, method = "euclidean")
```

### Computing hierarchical clustering

```{r}
hclust_ward <- hclust(proxmat, method = "ward.D")
plot(hclust_ward, cex = 0.6)
```

### Comparing clustering linkage methods

```{r}
m <- c("average", "single", "complete", "ward")
names(m) <- m

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

### Determining optimal number of clusters

```{r}
set.seed(1234)
gap_stat <- clusGap(shan_ict, FUN = hcut, nstart = 25, K.max = 10, B = 50)
fviz_gap_stat(gap_stat)
```

### Interpreting the dendrogram

```{r}
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, k = 6, border = 2:5)
```

### Visually-driven hierarchical clustering with heatmaply

```{r fig.width=10, fig.height=8}
heatmaply(normalize(shan_ict),
          Colv = NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          k_row = 6,
          margins = c(NA, 200, 60, NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main = "Geographic segmentation of Shan State by ICT indicators",
          xlab = "ICT indicators",
          ylab = "Townships")
```

### Mapping the clusters

```{r}
groups <- as.factor(cutree(hclust_ward, k = 6))

shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(CLUSTER = `as.matrix.groups.`)
```

```{r fig.width=10}
qtm(shan_sf_cluster, "CLUSTER")
```

## Spatially constrained clustering: SKATER method

### Converting to SpatialPolygonsDataFrame

```{r}
shan_sp <- as_Spatial(shan_sf)
```

### Computing neighbour list

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

```{r fig.width=10}
plot(st_geometry(shan_sf), border = grey(.5))
pts <- st_coordinates(st_centroid(shan_sf))
plot(shan.nb, pts, col = "blue", add = TRUE)
```

### Computing minimum spanning tree

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
```

```{r}
shan.w <- nb2listw(shan.nb, lcosts, style = "B")
summary(shan.w)
```

```{r}
shan.mst <- mstree(shan.w)
```

```{r fig.width=10}
plot(st_geometry(shan_sf), border = gray(.5))
plot(shan.mst, pts, col = "blue", cex.lab = 0.7, cex.circles = 0.005, add = TRUE)
```

### Computing spatially constrained clusters using SKATER

```{r}
clust6 <- skater(edges = shan.mst[, 1:2],
                  data = shan_ict,
                  method = "euclidean",
                  ncuts = 5)
```

```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(SP_CLUSTER = `as.factor.groups_mat.`)
```

```{r fig.width=10}
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

## Spatially constrained clustering: ClustGeo method

### Non-spatially constrained hierarchical clustering

```{r}
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, k = 6, border = 2:5)
```

```{r}
groups <- as.factor(cutree(nongeo_cluster, k = 6))
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(CLUSTER = `as.matrix.groups.`)
```

```{r fig.width=10}
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

### Spatially constrained hierarchical clustering

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

```{r}
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K = 6, graph = TRUE)
```

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)

groups <- as.factor(cutree(clustG, k = 6))
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(CLUSTER = `as.matrix.groups.`)
```

```{r fig.width=10}
qtm(shan_sf_Gcluster, "CLUSTER")
```

## Visual interpretation of clusters

```{r}
ggplot(data = shan_sf_ngeo_cluster, aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

```{r fig.width=10}
keep_vars <- c("RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")

ggparcoord(data = shan_sf_ngeo_cluster,
           columns = match(keep_vars, names(shan_sf_ngeo_cluster)),
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Multiple parallel coordinates plot of ICT variables by cluster") +
  facet_wrap(~ CLUSTER) +
  theme(axis.text.x = element_text(angle = 30))
```
