---
title: "Hands-on Exercise 3: Point Pattern Analysis"
author: "Andrew Huang"
date: "2026-09-04"
date-modified: "last-modified"
format: html
execute:
  warning: false
  freeze: auto
---

## Getting started

Ex01 and Ex02 were both basically about mapping — where things are, how to make the map look decent. This one's different: are these points actually clustered, or does it just look that way to the eye? Reusing the same two files from Ex01, `PreSchoolsLocation.geojson` and `MasterPlan2014SubzoneBoundaryNoSea.geojson`, except the subzone file isn't for drawing an administrative map this time, it's just standing in for "the boundary of Singapore."

New package: `spatstat`. It doesn't understand `sf` objects at all, it wants its own format called `ppp`. So a good chunk of what follows is just converting things into a shape spatstat will actually accept.

```{r}
pacman::p_load(sf, spatstat, tmap, tidyverse)
```

## Importing and preparing the data

```{r}
preschool <- st_read("data/geospatial/PreSchoolsLocation.geojson") %>%
  st_transform(crs = 3414)

mpsz <- st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson") %>%
  st_transform(crs = 3414)
```

The textbook version uses a proper coastline shapefile as the study area. Don't have that file, so instead all 323 subzone polygons get dissolved into one shape with `st_union()` — same land area in the end, just built a different way.

```{r}
sg_sf <- mpsz %>%
  st_union() %>%
  st_sf()

plot(st_geometry(sg_sf))
```

## Converting to spatstat's ppp format

`spatstat` wants two things: a `ppp` for the points, an `owin` for the window they live in. Sounds simple enough, except the dissolved outline from earlier still carries every single vertex from all 323 original polygons — thousands of points just to describe a coastline. Didn't think this would matter until later steps just sat there not finishing. Simplifying the outline first with `st_simplify()` (20m tolerance, invisible at Singapore's scale) fixed it — several of the point pattern functions later do real geometric work against the window boundary, and an unnecessarily detailed polygon makes all of that painfully slow.

```{r}
sg_sf_simplified <- sg_sf %>%
  st_simplify(dTolerance = 20, preserveTopology = TRUE)

sg_owin <- as.owin(sg_sf_simplified)

preschool_ppp <- as.ppp(st_coordinates(preschool), W = sg_owin)

plot(preschool_ppp)
```

```{r}
summary(preschool_ppp)
```

Two preschools can land on the exact same coordinate if they share a building, and duplicate points break some of the tests below, so I checked before going further:

```{r}
any(duplicated(preschool_ppp))
```

Turned out to be `TRUE` here, so the duplicates get jittered apart with `rjitter()` — nudges them a tiny random distance so they're not stacked exactly on top of each other:

```{r}
preschool_ppp_jit <- rjitter(preschool_ppp, retry = TRUE, nsim = 1, drop = TRUE)
any(duplicated(preschool_ppp_jit))
```

## Part 1 — First-order analysis: where are preschools concentrated?

First-order analysis is really just: where's it denser, where's it sparser. The tool is kernel density estimation (KDE) — each point gets smeared into a small "bump," and stacking all the bumps together gives a continuous surface instead of a scatter of dots.

### Fixed bandwidth

`sigma` sets how wide each bump is. Starting with a fixed number just to see what it looks like.

```{r}
kde_fixed <- density(preschool_ppp_jit,
                      sigma = 500,
                      edge = TRUE,
                      kernel = "gaussian")
plot(kde_fixed, main = "Fixed bandwidth (sigma = 500m)")
```

### Automatic bandwidth selection

500 was picked out of thin air, honestly. `spatstat` has actual selectors that work out a bandwidth from the data instead of guessing:

```{r}
bw.diggle(preschool_ppp_jit)
bw.CvL(preschool_ppp_jit)
bw.scott(preschool_ppp_jit)
bw.ppl(preschool_ppp_jit)
```

They don't agree with each other, which makes sense once you notice they're optimising for different things — `bw.diggle` leans toward picking up clusters, `bw.CvL`/`bw.scott` aim more for a smooth overall estimate. Plotting two of them side by side shows how much the bandwidth choice actually changes what you see:

```{r}
kde_diggle <- density(preschool_ppp_jit, sigma = bw.diggle, edge = TRUE)
kde_ppl <- density(preschool_ppp_jit, sigma = bw.ppl, edge = TRUE)

par(mfrow = c(1, 2))
plot(kde_diggle, main = "bw.diggle")
plot(kde_ppl, main = "bw.ppl")
par(mfrow = c(1, 1))
```

### Clark-Evans test

KDE gives you a picture, not a number. Clark-Evans actually gives a ratio and a p-value — compares the real average nearest-neighbour distance to what you'd expect if the points were scattered completely at random (CSR). Ratio under 1 means more clustered than random, over 1 means more spread out than random.

```{r}
clarkevans.test(preschool_ppp_jit,
                 correction = "none",
                 clipregion = sg_owin,
                 alternative = "clustered",
                 nsim = 99)
```

## Part 2 — Second-order analysis: is the clustering statistically real?

The KDE stuff above shows where it's dense, but doesn't really prove anything on its own — a cluster could just be noise. Second-order analysis instead looks at every pair of points and asks, at a given distance, is there more or less company than you'd expect from pure randomness?

Same recipe for all four functions below: compute the real statistic, simulate a bunch of random point patterns in the same window (Monte Carlo), and check whether the real line sits outside the band those random simulations produce.

### G-function — nearest neighbour distance

```{r}
G <- Gest(preschool_ppp_jit)
plot(G, main = "G-function")
```

```{r}
G.csr <- envelope(preschool_ppp_jit, Gest, nsim = 39)
plot(G.csr, main = "G-function vs CSR envelope")
```

### F-function — empty space distance

```{r}
F <- Fest(preschool_ppp_jit)
plot(F, main = "F-function")
```

```{r}
F.csr <- envelope(preschool_ppp_jit, Fest, nsim = 39)
plot(F.csr, main = "F-function vs CSR envelope")
```

### K-function — points within a given radius

This one actually got stuck the first time round. K-function checks every pair of points against every other, already heavier than G or F, and spatstat's default edge correction (isotropic) does extra geometry against the window boundary on top of that — with Singapore's mainland-plus-a-bunch-of-small-islands shape, it basically never finished. Switched to border correction instead: simpler, much faster, still a valid estimate.

```{r}
K <- Kest(preschool_ppp_jit, correction = "border")
plot(K, . - r ~ r, main = "K-function")
```

```{r}
K.csr <- envelope(preschool_ppp_jit, Kest, nsim = 39, correction = "border")
plot(K.csr, . - r ~ r, main = "K-function vs CSR envelope")
```

### L-function — a rescaled, easier-to-read version of K

```{r}
L <- Lest(preschool_ppp_jit, correction = "border")
plot(L, . - r ~ r, main = "L-function")
```

```{r}
L.csr <- envelope(preschool_ppp_jit, Lest, nsim = 39, correction = "border")
plot(L.csr, . - r ~ r, main = "L-function vs CSR envelope")
```

## Reading the envelope plots

The grey band in each plot is the range you'd expect from 39 random simulations in the same window. Where the black (observed) line pokes above the band, points are clustering at that distance; where it dips below, they're more spread out than random; where it just sits inside the band, can't really tell the pattern apart from CSR at that scale.

## Notes and open questions

This still isn't the textbook's actual dataset — it's Ex01's preschool and subzone files again, standing in for the childcare-centre data the chapter uses, and the study boundary is a dissolved subzone outline rather than the real coastline shapefile. Same substitution as before, just flagging it again here since it affects every result in this file.

Testing CSR against the whole island is a pretty blunt way to ask the question, too — preschools are obviously going to cluster wherever people actually live, so a chunk of what these tests are picking up is probably just population density, not anything specific to where a preschool gets built. Splitting this by planning area and re-running would be a better test of that, if there's time before this needs to be handed in.
