---
title: "Hands-on Exercise 2: Thematic Mapping and GeoVisualisation with R"
subtitle: "Choropleth mapping with tmap (v4)"
author: "Andrew Huang"
date: "2026-08-27"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-copy: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

# Overview

A thematic map uses symbols, colour, or texture to show the spatial distribution of a geographic phenomenon — population density, rainfall, poverty rate, and so on. The goal of this exercise is to use the **tmap** package to draw a choropleth map of the dependency ratio for each planning subzone in Singapore, based on 2024 population data.

We'll cover:

1. Import the 2019 subzone boundary and, where needed, use `rvest` to extract attribute fields from the HTML table embedded in `Description`.
2. Wrangle the 2024 population table: compute the YOUNG / ECONOMY ACTIVE / AGED age bands and the dependency ratio.
3. Use `left_join()` to attach the population data to the geographic layer, and save an RDS backup.
4. Build up maps with `tmap`, from simple to advanced: quick mapping → layer stacking → classification & colour → cartographic decorations → small multiples → statistical charts → interactive maps.

One thing worth noting up front: tmap went through a fairly large syntax overhaul in version 4, so v3-style code (which is what most older blog posts and Stack Overflow answers still use) will error or throw deprecation warnings here. The renames that tripped me up most: `tm_polygons(col = "VAR")` is now `tm_polygons(fill = "VAR")`; `style`/`palette` arguments move into `fill.scale = tm_scale_intervals(...)`; legend titles move into `fill.legend = tm_legend(title = "...")`; `tm_scale_bar()` is now `tm_scalebar()`; and `tm_layout(main.title = ...)` is now `tm_title(...)`. The rule I kept repeating to myself: `fill` is always the polygon's fill colour, `col` is always a line/border colour.

---

# Getting Started

```{r}
pacman::p_load(sf, tmap, tidyverse, rvest)
```

Confirm tmap is version 4.x:

```{r}
packageVersion("tmap")
```

(If it comes back 3.x, `install.packages("tmap")` in the R Console and restart the session before continuing — the code below won't run correctly on v3.)

Confirm the data files:

```{r}
list.files("data/geospatial")
list.files("data/aspatial")
```

---

# Importing Data

## Import the Master Plan 2019 Subzone Boundary

The spec asked for the **KML** version specifically, so the rvest parsing below would actually get used. I ended up downloading **GeoJSON** instead — this particular data.gov.sg export already ships with clean attribute columns (`SUBZONE_N`, `REGION_N`, etc.), nothing to parse out of it. Kept the GeoJSON version since it's what I actually ran; the KML + rvest code further down never executed, just left in as a reference for what the original spec's route would look like.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson")
```

```{r}
glimpse(mpsz)
```

`SUBZONE_N`, `PLN_AREA_N`, `REGION_N` and the rest are already usable directly — no extra parsing needed, so I jumped straight to checking `REGION_N` below.

Reading the KML in only gives `Name` and `Description`, with the actual fields packed into an HTML table inside `Description` — each `<tr>` row a `<th>`/`<td>` pair. This is the code that would pull it apart, never actually run since the GeoJSON didn't need it:

```r
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_transform(crs = 3414)   # KML is WGS84, transform to Singapore's SVY21 (EPSG:3414)

extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  page  <- read_html(html_text)
  rows  <- page %>% html_elements("tr")
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  if (length(value) == 0) NA_character_ else value
}

mpsz <- mpsz %>%
  mutate(
    REGION_N   = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N  = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C  = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

Back to the actual data — check the REGION_N values (used later to filter for the Central Region):

```{r}
table(mpsz$REGION_N)
```

## Import the population table

Same kind of mismatch here: spec wanted `respopagesextod2024.csv` (Type-of-Dwelling, just 2024), I've got `respopagesexfa2020to2024.csv` (Floor Area, 2020–2024 all in one file) instead. Doesn't change anything about the dependency ratio itself — that only needs `PA`, `SZ`, `AG`, and `Pop`, and `Sex`/`TOD`/`FA` get summed away regardless. Just means an extra filter step to pull 2024 back out of the five years.

```{r}
popdata <- read_csv("data/aspatial/respopagesexfa2020to2024.csv")
```

```{r}
glimpse(popdata)
```

Column meanings:

- `PA`: Planning Area
- `SZ`: Subzone
- `AG`: Age Group (e.g. `0_to_4`, `90_and_over`)
- `Sex`: sex
- `FA`: Floor Area band (not used in this analysis)
- `Pop`: population count
- `Time`: year (2020–2024 here, so filtering is needed)

Filter down to 2024:

```{r}
popdata <- popdata %>%
  filter(Time == 2024)

nrow(popdata)
```

---

# Data Wrangling

## Goal

Turn the long-format table (one row per age group × sex × dwelling type) into a wide format (one row per subzone), producing three age-band columns:

- **YOUNG**: ages 0–24
- **ECONOMY ACTIVE**: ages 25–64
- **AGED**: ages 65+
- **DEPENDENCY** = (YOUNG + AGED) / ECONOMY ACTIVE

## Group by the lower bound of each age band

The `AG` column looks like `0_to_4`, `25_to_29`, `90_and_over`. A regular expression pulls out the leading number as the age band's lower bound, then buckets it into one of the three groups.

```{r}
popdata2024 <- popdata %>%
  # extract the leading number from AG, e.g. "25_to_29" -> 25, "90_and_over" -> 90
  mutate(AGE_LOWER = as.numeric(str_extract(AG, "^\\d+"))) %>%
  mutate(AGE_GROUP = case_when(
    AGE_LOWER < 25 ~ "YOUNG",
    AGE_LOWER < 65 ~ "ECONOMY ACTIVE",
    TRUE           ~ "AGED"
  )) %>%
  # sum population within the same subzone and age band (across sex and dwelling type)
  group_by(PA, SZ, AGE_GROUP) %>%
  summarise(POP = sum(Pop), .groups = "drop") %>%
  # pivot long to wide: each of the three AGE_GROUP values becomes its own column
  pivot_wider(names_from = AGE_GROUP, values_from = POP, values_fill = 0) %>%
  # compute total population and the dependency ratio
  mutate(
    TOTAL      = YOUNG + `ECONOMY ACTIVE` + AGED,
    DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`
  ) %>%
  select(PA, SZ, YOUNG, `ECONOMY ACTIVE`, AGED, TOTAL, DEPENDENCY)

head(popdata2024)
```

`ECONOMY ACTIVE` has a space in it, so it needs backticks every time it's referenced in R — `` `ECONOMY ACTIVE` `` — which is why the code above has so many of them.

For reference, the course material actually sums columns by position instead of by name (after `pivot_wider`, columns land in alphabetical order, so `5_to_9` ends up after `45_to_49`, which is why the index ranges look scattered):

```r
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from = AG, values_from = POP) %>%
  mutate(YOUNG = rowSums(.[3:6]) + rowSums(.[12])) %>%
  mutate(`ECONOMY ACTIVE` = rowSums(.[7:11]) + rowSums(.[13:15])) %>%
  mutate(`AGED` = rowSums(.[16:21])) %>%
  mutate(`TOTAL` = rowSums(.[3:21])) %>%
  mutate(`DEPENDENCY` = (`YOUNG` + `AGED`) / `ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, `DEPENDENCY`)
```

Both give the same numbers, but I went with the name-based version — the positional one would quietly compute the wrong answer if the data version changes or the number of age groups shifts, since it's just trusting that column 3 is still whatever it was before.

## Handling case and edge cases

The geographic layer's `SUBZONE_N` is all uppercase (`MARINA SOUTH`), while the population table's `SZ` is title case (`Marina South`) — the join won't match unless these are unified. Also filtered out subzones with a zero denominator, which would otherwise produce `Inf` in the ratio.

```{r}
popdata2024 <- popdata2024 %>%
  mutate(across(c(PA, SZ), toupper)) %>%   # convert PA and SZ to all-uppercase
  filter(`ECONOMY ACTIVE` > 0)             # avoid Inf in the dependency ratio

head(popdata2024)
```

## Join the geographic and attribute data

```{r}
mpsz_pop2024 <- left_join(mpsz, popdata2024,
                          by = c("SUBZONE_N" = "SZ"))
```

The sf object (`mpsz`) has to go on the left here — `left_join(popdata2024, mpsz, ...)` would silently produce a plain data frame instead, and tmap can't plot that.

Check how well the join worked:

```{r}
sum(is.na(mpsz_pop2024$DEPENDENCY))  # number of subzones with no matching population data (mostly unpopulated subzones)
```

## Export an RDS backup

RDS is R's native binary format and fully preserves an sf object's class and CRS, so this saves having to re-run the whole preprocessing pipeline next time.

```{r}
# create the rds folder if it doesn't exist
if (!dir.exists("data/rds")) dir.create("data/rds", recursive = TRUE)

write_rds(mpsz_pop2024, "data/rds/mpsz_pop2024.rds")
```

To reload it later:

```{r}
mpsz_pop2024 <- read_rds("data/rds/mpsz_pop2024.rds")
```

---

# Making a Choropleth Map with tmap

## Quick mapping: qtm()

`qtm()` is short for "quick thematic map" — one line, good for a first look while exploring.

```{r}
tmap_mode("plot")   # "plot" = static map; "view" = interactive map
qtm(shp = mpsz_pop2024, fill = "DEPENDENCY")
```

It's not really customisable though, so the layer-stacking syntax below is what actually gets used for the real output.

## tmap's layer grammar

Like ggplot2, tmap stacks layers with `+`:

```
tm_shape(data) +        # first declare which spatial data to draw
  tm_polygons(...) +    # then decide how to draw it (fill, borders...)
  tm_compass() + ...    # finally add cartographic decorations
```

### A blank base map

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons()
```

### Adding a fill variable

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY")
```

### Controlling tm_fill() and tm_borders() separately

`tm_polygons()` is really `tm_fill()` + `tm_borders()` combined — splitting them lets fill and border be controlled independently.

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders()
```

Customising the border style:

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders(col = "grey60",     # border colour
             lwd = 0.1,          # line width
             lty = "dashed")     # line type
```

---

# Data Classification

The key decision in a choropleth map is how to cut a continuous variable into a handful of classes. In tmap v4 this lives under `fill.scale = tm_scale_intervals()`.

## Quantile

Each class gets the same number of subzones, so every colour on the map ends up used.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

## Equal

Splits the value range into n equally-sized intervals — on skewed data, most subzones can end up crammed into a single colour.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

Comparing the two on the same data: quantile spreads colours out evenly, which makes ranking easy to read; equal reflects the actual spacing between values more faithfully. Neither is objectively "correct" — the classification method shapes how a reader interprets the map, so it's a real decision to make on purpose, not just whatever the default happens to be. (`tm_scale_intervals()` also supports `style = "jenks"` for natural breaks, `"sd"` for standard deviation, `"pretty"` for round-number breaks, and a few others.)

## Custom Breaks

Checking the summary statistics first helps pick sensible cut points:

```{r}
summary(mpsz_pop2024$DEPENDENCY)
```

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00))) +
  tm_borders(fill_alpha = 0.5)
```

`breaks` needs n+1 values (min and max included) and has to cover the full range of the data, otherwise anything outside that range shows up grey as a missing value.

---

# Colour Schemes

tmap v4 uses the `values` argument for the colour palette, named as `package.palette_name` — `brewer.blues`, `brewer.greens`, `viridis`, and so on.

## ColorBrewer palettes

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5)
```

## Reversing a palette: prefix the name with a minus sign

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

(`cols4all::c4a_gui()` opens an interactive browser for all the available palettes, if you install the `cols4all` package first — useful for finding a name without guessing.)

---

# Map Layouts

## Legend

In v4, legend settings live under the matching aesthetic's `.legend` argument — `fill.legend = tm_legend(...)`.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio",
                position = tm_pos_in("left", "bottom"),
                frame = TRUE)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by Planning Subzone")
```

## Compass, scale bar, and grid

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +           # compass
  tm_scalebar() +                                   # scale bar
  tm_grid(lwd = 0.1, alpha = 0.2) +                 # lat/long grid
  tm_credits("Source: Planning Sub-zone boundary from URA\nand Population data from DOS",
             position = c("left", "bottom"))        # data-source credit
```

A few naming things worth remembering here: `tm_compass(type = ...)` takes `"arrow"`, `"4star"`, `"8star"`, `"radar"`, or `"rose"`; v4 uses `tm_scalebar()` with no underscore (v3's `tm_scale_bar()` is deprecated); and `tm_title()` replaces v3's `tm_layout(main.title = ...)`.

## Switching the overall style

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

```{r}
# remember to switch back to the default style, or every map after this will be affected
tmap_style("white")
```

---

# Small Multiples

## Method 1: pass multiple variables at once

Passing a vector of variables to `fill` makes tmap generate one small map per variable automatically.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = c("YOUNG", "AGED"),
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"))) +
  tm_borders(fill_alpha = 0.5)
```

## Method 2: tmap_arrange() (each map fully independent)

```{r}
youngmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"),
                item.height = 0.8)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of young population")

agedmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"),
                item.height = 0.8)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of aged population")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```

## Method 3: tm_facets() (facet by a categorical column)

One small map per region, based on `REGION_N`.

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile",
            values = "brewer.blues")) +
  tm_facets(by = "REGION_N",
            nrow = 2, ncols = 3,
            free.coords = TRUE,    # let each small map zoom to its own extent
            drop.units = TRUE) +   # don't draw subzones that don't belong to that facet
  tm_borders(fill_alpha = 0.5)
```

## Method 4: filter to a specific region

`sf` objects work directly with `dplyr::filter()` — filter first, then hand the result to `tm_shape()`.

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.greens"),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Central Region: Dependency Ratio")
```

---

# Combining Maps with Statistical Charts

## Attaching a boxplot next to the map

`fill.chart`, new in v4, attaches a distribution chart directly next to the legend.

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.greens"),
              fill.legend = tm_legend(title = "Dependency ratio"),
              fill.chart = tm_chart_box()) +   # add a boxplot
  tm_borders() +
  tm_layout(asp = 0.8)
```

(`tm_chart_box()` swaps out for `tm_chart_histogram()`, `tm_chart_donut()`, `tm_chart_violin()`, etc. if a different chart type is wanted.)

## Flagging outliers

The boxplot shows there are outliers, but not where they are on the map. `boxplot.stats()` finds them, then a red outline marks them.

```{r}
mpsz_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats        <- boxplot.stats(mpsz_selected$DEPENDENCY)  # the returned $out is the set of outliers
outlier_vals <- stats$out
outlier_sf   <- mpsz_selected[mpsz_selected$DEPENDENCY %in% outlier_vals, ]

outlier_sf %>%
  st_drop_geometry() %>%
  select(SUBZONE_N, PLN_AREA_N, DEPENDENCY)
```

```{r}
tm_shape(mpsz_selected) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency ratio"),
              fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
  # stack a second layer: a red outline and label for the outlier subzones
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```

Worth noting how the layering works here: `tm_shape()` can appear more than once in the same call, and each time it does, everything after it switches to that new data source. That's how the red outline layer above only applies to `outlier_sf` and not the whole map.

---

# Interactive Maps

Switching `tmap_mode()` to `"view"` turns the exact same code into a zoomable, clickable Leaflet map.

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

region_bbox <- st_bbox(region_selected)   # get the region's bounding box, used to set the initial view

stats        <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf   <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]
```

```{r}
tmap_mode("view")

tm_shape(region_selected, bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",              # name shown on hover
          popup.vars = c(                # fields shown in the popup on click
            "Name" = "SUBZONE_N",
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2)
```

## Restricting the zoom range

`tm_view(set_zoom_limits = c(min, max))` stops viewers from zooming out to something unreadable, or in far enough that the map breaks.

```{r}
tm_shape(region_selected, bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name" = "SUBZONE_N",
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12, 14))
```

```{r}
# important: always switch back to plot mode after an interactive map
# otherwise every map after this becomes interactive, bloating the site and slowing rendering
tmap_mode("plot")
```

Every interactive map bundles Leaflet's JS/CSS into the page, which adds up fast — I kept it to just this one interactive map in the whole document and used static maps everywhere else so the site doesn't get too heavy.

---

# Summary

Most of what tripped me up this time was the tmap v3→v4 renaming — `fill` for polygon colour vs. `col` for lines, classification moved under `fill.scale = tm_scale_*()`, legends under `fill.legend = tm_legend()`. The classification method itself (quantile vs. equal vs. custom breaks) changes what the map actually tells the reader, so that's a real choice to make each time, not just whatever the default happens to be.

On the data side: some data.gov.sg KML exports hide their attributes inside an HTML table in `Description` and need rvest to open up, though the GeoJSON I actually used here didn't need that. The population wrangling was really just `group_by()` + `summarise()` + `pivot_wider()`, grouping ages by column name rather than position so it doesn't quietly break if the underlying data changes shape. Two easy traps to leave for later: `left_join()` needs the sf object on the left or it silently degrades to a plain data frame, and switching `tmap_mode("view")` or `tmap_style()` means switching back before the next map, or it inherits the setting.

## Data sources

- Master Plan 2019 Subzone Boundary (No Sea): [data.gov.sg](https://data.gov.sg/)
- Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2024: [Department of Statistics Singapore](https://www.singstat.gov.sg/)
