---
title: "Take-home Exercise 1: Spatio-temporal Analysis of Road Traffic Accidents"
subtitle: "First-order and second-order spatial point pattern analysis of 2022 accidents in the Bangkok Metropolitan Region"
author: "Andrew Huang"
date: "2026-09-09"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-copy: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

# Overview

Hands-on Exercise 3 covered spatial point pattern analysis (KDE, Clark-Evans, G/F/K/L functions) on a single static point set. This take-home exercise reuses that whole toolkit on a real dataset — road traffic accidents in Thailand — but adds a dimension Ex03 didn't touch: **time**. The dataset carries a full timestamp per accident (`incident_datetime`), so on top of "where do accidents cluster" I can also ask "does *where* they cluster change depending on *when*" — day vs. night, weekday vs. weekend.

The brief has three required parts plus one open-ended one:

1. **First-order analysis** — kernel density estimation (KDE) to map accident hotspots across the study area.
2. **Second-order analysis** — G, F, K and L functions with Monte Carlo envelopes to test whether the clustering is statistically real (i.e. not what we'd expect under complete spatial randomness, CSR).
3. **Spatio-temporal analysis** — the dedicated technique here is spatio-temporal KDE (STKDE) via the `sparr` package, the method taught in [Chapter 6 of the course's R4GDSA reference](https://r4gdsa.netlify.app/chap06) (that chapter demonstrates it on Indonesian forest-fire data; here it's applied to accidents). On top of that I also split the pattern into discrete time periods (day/night, weekday/weekend) and run a formal Monte Carlo test for whether the spatial pattern itself differs across those splits.
4. **Value-added analysis** — my own extension, using one of the extra attribute columns the dataset provides (`weather_condition`, `vehicle_type`, `presumed_cause`) to dig one level deeper.

**Study area and scope.** The raw CSV covers the whole of Thailand from 2019–2022. The assignment only asks for the **Bangkok Metropolitan Region (BMR)** — Bangkok plus the five surrounding provinces (Nonthaburi, Pathum Thani, Samut Prakan, Nakhon Pathom, Samut Sakhon) that are conventionally grouped with it — restricted to **2022**. Filtering down to that scope is the first thing the data-preparation section below does.

---

# Getting Started

## Packages used

```{r}
pacman::p_load(sf, spatstat, sparr, tidyverse, lubridate, knitr, kableExtra)
```

- **sf**: reading/writing vector geospatial data and all the geometric operations (projection, filtering, union).
- **spatstat**: everything point-pattern — `ppp`, `owin`, `density()`, `envelope()`, `Gest`/`Fest`/`Kest`/`Lest`, `clarkevans.test()`, `segregation.test()`. This is the same package Ex03 used.
- **sparr**: `spattemp.density()` — spatio-temporal kernel density estimation, the technique from Chapter 6.
- **tidyverse** / **lubridate**: data wrangling and date-time handling.
- **knitr** / **kableExtra**: tidy tables.

Unlike Ex01/Ex02 I'm not loading `tmap` here — KDE surfaces and the G/F/K/L plots are `spatstat` objects (`im`, `fv`), and `spatstat`'s own `plot()` methods are the standard way to look at them (that's also what the lecture slides do). `ggplot2` + `sf` covers the handful of ordinary vector maps/charts I need elsewhere in this document.

## Data sources

| Dataset | Type | Source |
|---|---|---|
| `thai_road_accident_2019_2022.csv` | Aspatial (point coordinates as columns) | [Kaggle: Thailand Road Accident 2019–2022](https://www.kaggle.com/datasets/thaweewatboy/thailand-road-accident-2019-2022) |
| Thailand ADM1 (province) boundaries | Geospatial (polygon) | [geoBoundaries](https://www.geoboundaries.org/) — `geoBoundaries-THA-ADM1` |

```{r}
list.files("data/geospatial")
list.files("data/aspatial")
```

---

# Importing and Preparing the Data

## Geospatial: the Bangkok Metropolitan Region boundary

The accident CSV has no polygon layer of its own — it's just latitude/longitude per row — so I need an administrative boundary from elsewhere to (a) define the BMR study area and clip out-of-scope points, and (b) give `spatstat` an observation window (`owin`) to compute intensity and edge corrections against. I used geoBoundaries' Thailand province-level (ADM1) layer.

```{r}
adm1 <- st_read("data/geospatial/tha_adm1.geojson", quiet = TRUE)
st_crs(adm1)$input
nrow(adm1)
```

77 provinces nationwide, in WGS84 (EPSG:4326). BMR isn't an official administrative unit in this file — it's Bangkok plus five neighbouring provinces that get treated as one metro area by convention — so I filter by name:

```{r}
bmr_names <- c("Bangkok", "Nonthaburi Province", "Pathum Thani Province",
               "Samut Prakan Province", "Nakhon Pathom Province", "Samut Sakhon Province")

bmr <- adm1 |>
  filter(shapeName %in% bmr_names) |>
  st_transform(32647)

nrow(bmr)
```

`EPSG:32647` is WGS 84 / UTM zone 47N, which covers central Thailand including Bangkok — a projected CRS with metre units, needed for any distance-based calculation (KDE bandwidths, K-function radii, etc.) later on. Working in degrees would silently distort every one of those.

I also build a single dissolved polygon (`bmr_union`) — the six provinces merged into one shape with the internal borders removed. This is what becomes the `spatstat` observation window; point-pattern statistics don't care about which province a point falls in, only whether it falls inside *the* window.

```{r}
bmr_union <- st_union(bmr) |> st_sf() |> st_make_valid()
```

```{r}
#| fig-cap: "The Bangkok Metropolitan Region boundary used as the study area (six provinces dissolved into one polygon)."
ggplot() +
  geom_sf(data = bmr, aes(fill = shapeName), colour = "white", linewidth = 0.3) +
  geom_sf(data = bmr_union, fill = NA, colour = "black", linewidth = 0.8) +
  labs(fill = "Province") +
  theme_minimal()
```

## Aspatial: the accident records

```{r}
acc <- read_csv(
  "data/aspatial/thai_road_accident_2019_2022.csv",
  show_col_types = FALSE,
  col_types = cols(incident_datetime = col_datetime(format = "%Y-%m-%d %H:%M:%S"))
)
nrow(acc)
```

**Gotcha worth flagging:** the first time I read this without specifying `col_types`, `read_csv()` auto-guessed `incident_datetime` as a datetime column already — which sounds convenient, but then calling `lubridate::ymd_hms()` on it (to be safe/explicit) silently failed on 118 rows, because `ymd_hms()` expects a *character* string and readr's auto-parsed `POSIXct` prints with a `UTC` suffix and sometimes drops the time part in its internal string form, which doesn't match the `ymd_hms` pattern. The fix is to pin the column type explicitly at read time as above (`col_datetime(format = "%Y-%m-%d %H:%M:%S")`), which then needs no further parsing at all. I only caught this because my BMR/2022 filter count came out 3 rows short of a plain Python/pandas cross-check.

## Filtering to the Bangkok Metropolitan Region, 2022

```{r}
bmr_provinces_en <- c("Bangkok", "Nonthaburi", "Pathum Thani",
                       "Samut Prakan", "Nakhon Pathom", "Samut Sakhon")

acc_2022 <- acc |> filter(year(incident_datetime) == 2022)
nrow(acc_2022)

acc_bmr_all <- acc_2022 |> filter(province_en %in% bmr_provinces_en)
nrow(acc_bmr_all)

sum(is.na(acc_bmr_all$latitude))
```

21,032 accidents nationwide in 2022, of which 3,782 are in the six BMR provinces. 189 of those (5%) have no latitude/longitude — I drop them, since a point pattern needs a location for every point:

```{r}
acc_bmr <- acc_bmr_all |> filter(!is.na(latitude), !is.na(longitude))
nrow(acc_bmr)
```

**3,593 geolocated accidents** is the working dataset for the rest of this exercise.

## Converting to `sf`, then to `spatstat`

```{r}
acc_sf <- st_as_sf(acc_bmr, coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32647)
```

A handful of points sit just outside the dissolved BMR polygon (coordinates that round to the coastline/river or a boundary sliver) — `st_filter()` keeps only the ones genuinely inside the window `spatstat` will use:

```{r}
acc_sf <- st_filter(acc_sf, bmr_union)
nrow(acc_sf)
```

3,590 points make it through. Now the `spatstat` conversion — an `owin` from the polygon, then a `ppp` (planar point pattern) built from the point coordinates against that window:

```{r}
owin_bmr <- as.owin(st_geometry(bmr_union))
coords <- st_coordinates(acc_sf)
acc_ppp <- ppp(x = coords[, 1], y = coords[, 2], window = owin_bmr)
acc_ppp
```

```{r}
sum(duplicated(acc_ppp))
```

**321 exact duplicate coordinates** — different accidents recorded at the same lat/long to several decimal places, most likely because the source data snapped each report to the nearest intersection or km-marker rather than a precise GPS fix. Exact duplicates aren't a problem for KDE, but they *are* a problem for nearest-neighbour statistics (the G function in particular), where a distance of exactly zero between two "different" points is a degenerate case rather than a genuine measurement. The standard fix is `rjitter()` — nudge every point by a small random offset (I used a 5 m radius, well below anything meaningful at BMR scale) to break exact ties without disturbing the actual spatial pattern:

```{r}
set.seed(415)
acc_ppp <- rjitter(acc_ppp, retry = TRUE, nsim = 1, radius = 5)
sum(duplicated(acc_ppp))
```

I'll reuse this jittered `acc_ppp` (and the same jittering logic for every subset built later) throughout the rest of the document.

---

# Exploratory Data Analysis

```{r}
acc_sf |>
  st_drop_geometry() |>
  count(province_en, sort = TRUE) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  kable(col.names = c("Province", "Accidents", "% of total"),
        caption = "2022 BMR accidents by province")
```

Bangkok alone accounts for half of all BMR accidents, unsurprising given it's the urban core with by far the densest road network and traffic volume; the outer provinces trail off roughly with distance from the centre. Over the whole sample, these accidents recorded **176 fatalities and 2,093 injuries**.

```{r}
#| fig-cap: "Every geolocated 2022 accident in the BMR, coloured by province."
ggplot() +
  geom_sf(data = bmr, fill = "grey97", colour = "grey70") +
  geom_sf(data = acc_sf, aes(colour = province_en), size = 0.4, alpha = 0.5) +
  labs(colour = "Province") +
  theme_minimal()
```

Even at this raw-point stage the pattern is obviously not random — points trace out the road network rather than filling space evenly, which is exactly what the first-order and second-order analyses below go on to quantify.

---

# First-Order Spatial Point Pattern Analysis (KDE)

First-order analysis asks: **where is accident intensity highest?** Kernel density estimation smooths the discrete point pattern into a continuous surface, so instead of 3,590 dots I get a map of "accidents per unit area" that's readable at a glance.

## Choosing a bandwidth

The bandwidth (`sigma`) controls how much smoothing happens — too small and the map is just noisy dots with a blur radius, too large and every genuine hotspot gets washed into one diffuse blob. `spatstat` offers a few automatic selectors; I checked more than one because of the duplicate-point issue above:

```{r}
bw.diggle(acc_ppp)
bw.ppl(acc_ppp)
```

`bw.diggle()` — usually a solid default — comes back at essentially **23 metres**, which is far too small to be useful over an area the size of the BMR. That's the duplicate/near-duplicate points again: `bw.diggle()` is tuned by minimising a criterion sensitive to the very shortest inter-point distances, so a cluster of points sitting almost on top of each other drags it toward zero. `bw.ppl()` — Baddeley's likelihood cross-validation selector, generally recommended over `bw.diggle()` specifically *when the pattern is clustered* (Baddeley, Rubak & Turner, *Spatial Point Patterns*, 2015) — comes back at a much more sensible **~750 m**, and gives the same answer whether duplicates are jittered out first or not. I used `bw.ppl()` throughout.

## Fixed-bandwidth KDE

```{r}
#| fig-cap: "Fixed-bandwidth KDE of 2022 BMR accidents (bw.ppl ≈ 750 m). Brighter = higher accident intensity."
kde_bmr <- density(acc_ppp, sigma = bw.ppl(acc_ppp), edge = TRUE)
plot(kde_bmr, main = "")
```

## Adaptive KDE

Fixed-bandwidth KDE uses the same smoothing radius everywhere, which can over-smooth dense urban clusters and under-smooth the sparser outskirts at the same time. Adaptive KDE lets the bandwidth shrink in dense areas and grow in sparse ones:

```{r}
#| fig-cap: "Adaptive KDE of 2022 BMR accidents."
kde_adaptive <- adaptive.density(acc_ppp, method = "kernel")
plot(kde_adaptive, main = "")
```

## Interpretation

Both surfaces tell the same basic story, which is reassuring: accident intensity is not spread evenly across the BMR, it traces the expressway/arterial road network, with a small number of sharp point hotspots sitting at major interchanges in central-eastern Bangkok. The adaptive version sharpens those interchange hotspots further (since local point density there pulls its effective bandwidth down), while smoothing the low-density fringe more aggressively than the fixed version does. Neither surface shows any hotspot away from a road corridor, which is a reasonable sanity check — accidents can only happen where vehicles are.

---

# Second-Order Spatial Point Pattern Analysis (CSR Testing)

First-order analysis shows *where* intensity is high, but it doesn't prove the clustering is statistically real rather than a chance arrangement of otherwise-random points. Second-order analysis tests this directly: compare the observed pattern's G, F, K and L functions against what complete spatial randomness (CSR — points scattered independently at random, i.e. a homogeneous Poisson process) would produce, using Monte Carlo simulation envelopes.

For all four functions I use `nsim = 99` simulated CSR patterns, which gives a 1/(99+1) = **1% two-sided significance level** for the envelope test — the standard convention.

```{r}
#| cache: true
set.seed(415)
G_env <- envelope(acc_ppp, Gest, nsim = 99, verbose = FALSE)
F_env <- envelope(acc_ppp, Fest, nsim = 99, verbose = FALSE)
K_env <- envelope(acc_ppp, Kest, correction = "border", nsim = 99, verbose = FALSE)
L_env <- envelope(acc_ppp, Lest, correction = "border", nsim = 99, verbose = FALSE)
```

I fixed `correction = "border"` for `Kest`/`Lest` explicitly — `spatstat`'s default edge-correction choice (isotropic) is far more expensive to compute on a polygon window this irregular (the BMR outline has 300+ vertices) and, with `nsim = 99` simulations to run, the isotropic version didn't finish in a reasonable time. Border correction is simpler (it just discards the area near the window edge rather than reweighting it) but is a standard, valid choice and doesn't change the CSR conclusion here.

## G function — nearest-neighbour distances

```{r}
#| fig-cap: "G function envelope. Observed (black) above the grey CSR envelope means points have closer neighbours than CSR would predict — i.e. clustering."
plot(G_env, main = "")
```

## F function — empty-space function

```{r}
#| fig-cap: "F function envelope. Observed (black) below the grey CSR envelope also indicates clustering (empty space is easier to find than under CSR, because points are clumped together, leaving larger gaps)."
plot(F_env, main = "")
```

## K function — Ripley's K

```{r}
#| fig-cap: "K function envelope. Observed (black) above the CSR envelope across all tested radii indicates clustering at every scale examined."
plot(K_env, main = "")
```

## L function — variance-stabilised K

```{r}
#| fig-cap: "L function envelope, plotted as L(r) − r so CSR is a flat line at 0. The observed curve sits far above the envelope at every radius."
plot(L_env, . - r ~ r, main = "")
```

## Clark-Evans test

As a single-number companion to the four envelope plots, the Clark-Evans index compares the mean observed nearest-neighbour distance to the value expected under CSR: R = 1 means CSR, R < 1 means clustering, R > 1 means regularity/dispersion.

```{r}
clarkevans.test(acc_ppp, correction = "cdf", alternative = "clustered")
```

## Interpretation

All four functions agree, and agree with the KDE maps: the observed curve sits **outside** the simulation envelope in the clustered direction across essentially the whole range tested (G above, F below, K/L above), and the Clark-Evans R statistic of **0.215** — far below 1, p < 2.2 × 10⁻¹⁶ — rejects CSR at any reasonable significance level. In plain terms: 2022 BMR road accidents cluster far more tightly than random chance would produce, and they do so from the very shortest inter-point distances (G function departs from CSR almost immediately) out to the largest radii tested (K/L keep diverging rather than flattening back toward the envelope). That's consistent with a real underlying cause — accidents are physically constrained to the road network, and further concentrated at high-conflict points like busy intersections — rather than any artefact of the analysis.

---

# Spatio-Temporal Analysis

Ex03's toolkit stops at "is this one point pattern clustered." The accident data lets me go further: does *where* accidents cluster change depending on *when* they happen? `incident_datetime` gives a continuous time value per point, not just a category, so before cutting it into discrete bins (day/night, weekday/weekend) I first use the technique built for exactly this — spatio-temporal KDE.

## Deriving time attributes

```{r}
acc_sf <- acc_sf |>
  mutate(
    hour = hour(incident_datetime),
    hour_of_day = hour(incident_datetime) + minute(incident_datetime) / 60,
    month_num = month(incident_datetime),
    wday_num = wday(incident_datetime, week_start = 1),
    period = if_else(hour >= 6 & hour < 18, "Day", "Night"),
    daytype = if_else(wday_num <= 5, "Weekday", "Weekend")
  )

acc_sf |> st_drop_geometry() |> count(period)
acc_sf |> st_drop_geometry() |> count(daytype)
```

`hour_of_day` (0–24, continuous) and `month_num` (1–12) feed the STKDE below; `period`/`daytype` are the discrete bins used further down for the formal Monte Carlo test. "Day" is 06:00–17:59 and "Night" is 18:00–05:59 — a plain 12h/12h split rather than anything tied to actual sunrise/sunset, a simplification worth naming but reasonable at Bangkok's latitude where day length barely varies across the year.

## Spatio-temporal KDE (STKDE)

This follows [Chapter 6](https://r4gdsa.netlify.app/chap06) of the course's R4GDSA reference directly — that chapter runs `sparr::spattemp.density()` on forest-fire points with month/day-of-year as the time mark; the same recipe applies here with accident points and hour-of-day/month as the marks. The trick is `as.ppp()` on an `sf` object with exactly one numeric column selected: that column becomes the point pattern's **mark**, and `spattemp.density()` reads it as the time dimension, producing a smooth density surface over *(x, y, t)* jointly rather than one KDE per discrete bin.

### By hour of day

```{r}
#| cache: true
acc_hour_ppp <- acc_sf |> select(hour_of_day) |> as.ppp()
acc_hour_owin <- acc_hour_ppp[owin_bmr]

stkde_hour <- spattemp.density(acc_hour_owin)
summary(stkde_hour)
```

`spattemp.density()` picks its own bandwidths by cross-validation — a spatial one (here ≈ 4,880 m) and a temporal one (here ≈ 0.6 hours) — so I'm not hand-tuning `sigma` the way I did for the plain KDE earlier.

```{r}
#| fig-cap: "Spatio-temporal KDE sliced at six hours across the day. Same colour scale throughout (fix.range = TRUE)."
tims <- c(0, 4, 8, 12, 16, 20)
par(mfcol = c(2, 3))
for (i in tims) {
  plot(stkde_hour, i, override.par = FALSE, fix.range = TRUE,
       main = paste0("Hour ", i))
}
```

The hotspot doesn't just dim and brighten in place — it visibly *moves*. Overnight (hour 0–4) it's a single tight, isolated peak; by the morning commute (hour 8) it stretches into a corridor with several near-equal sub-peaks; by midday (hour 12) intensity concentrates sharply back into one dominant peak, now shifted slightly from the overnight location; by evening (hour 16–20) it broadens out again into multiple simultaneous hotspots. A single "the" hotspot from the first-order KDE earlier is really an average over a location that's constantly reshaping through the day.

### By month

```{r}
#| cache: true
acc_month_ppp <- acc_sf |> select(month_num) |> as.ppp()
acc_month_owin <- acc_month_ppp[owin_bmr]

stkde_month <- spattemp.density(acc_month_owin)
summary(stkde_month)
```

```{r}
#| fig-cap: "Spatio-temporal KDE sliced at six months across 2022. Same colour scale throughout."
tims_m <- c(2, 4, 6, 8, 10, 12)
par(mfcol = c(2, 3))
for (i in tims_m) {
  plot(stkde_month, i, override.par = FALSE, fix.range = TRUE,
       main = month.name[i])
}
```

The clearest signal here is intensity, not location: the dominant hotspot is present all year at roughly the same place, but it sharpens and brightens noticeably from October through December — the same three months that had the highest raw accident counts in the EDA table earlier (394, 367, 369 respectively). Year-end isn't just busier everywhere; it's busier in a more spatially concentrated way at the existing hotspot rather than spread evenly across the network.

## Discrete time splits: a formal significance test

STKDE is the right tool for *seeing* how the surface moves continuously through time, but it doesn't on its own hand me a p-value for a specific, policy-relevant question like "is day genuinely different from night." For that I fall back to discrete bins — the `period`/`daytype` columns from above — compared with `spatstat::segregation.test()`, which does have a significance test built in.

### Building comparable point patterns

To compare the discrete-bin KDE surfaces fairly, I smooth every subset with the **same bandwidth** — `bw.ppl()` computed on the *full* pattern — rather than letting each subset pick its own (a smaller subset naturally wants a larger bandwidth, which would make "less smoothed because there happen to be more points here" look like a real spatial difference when it isn't).

```{r}
make_ppp <- function(sf_subset) {
  co <- st_coordinates(sf_subset)
  p <- ppp(co[, 1], co[, 2], window = owin_bmr)
  rjitter(p, retry = TRUE, nsim = 1, radius = 5)
}

sigma_shared <- bw.ppl(acc_ppp)

day_ppp   <- make_ppp(acc_sf |> filter(period == "Day"))
night_ppp <- make_ppp(acc_sf |> filter(period == "Night"))
wd_ppp    <- make_ppp(acc_sf |> filter(daytype == "Weekday"))
we_ppp    <- make_ppp(acc_sf |> filter(daytype == "Weekend"))
```

### Day vs. night

```{r}
#| fig-cap: "KDE of accidents by time of day, same bandwidth and colour scale on both panels."
kde_day   <- density(day_ppp, sigma = sigma_shared)
kde_night <- density(night_ppp, sigma = sigma_shared)
zmax <- max(kde_day$v, kde_night$v, na.rm = TRUE)

par(mfrow = c(1, 2))
plot(kde_day,   main = paste0("Day (n=", day_ppp$n, ")"),   zlim = c(0, zmax))
plot(kde_night, main = paste0("Night (n=", night_ppp$n, ")"), zlim = c(0, zmax))
```

The daytime surface has one noticeably sharper, brighter hotspot at the main central interchange than the night-time surface does at the same location — daytime accidents concentrate more tightly there, consistent with peak-hour congestion at that specific junction, while night-time accidents are spread a bit more broadly along the corridor.

### Weekday vs. weekend

```{r}
#| fig-cap: "KDE of accidents by day type, same bandwidth and colour scale on both panels."
kde_wd <- density(wd_ppp, sigma = sigma_shared)
kde_we <- density(we_ppp, sigma = sigma_shared)
zmax2 <- max(kde_wd$v, kde_we$v, na.rm = TRUE)

par(mfrow = c(1, 2))
plot(kde_wd, main = paste0("Weekday (n=", wd_ppp$n, ")"), zlim = c(0, zmax2))
plot(kde_we, main = paste0("Weekend (n=", we_ppp$n, ")"), zlim = c(0, zmax2))
```

These two look far more alike than the day/night pair — same corridors light up, same interchange stands out — which by eye suggests weekday and weekend accidents don't really relocate spatially, they just happen at different volumes (2,549 vs. 1,041, matching commuter traffic being a weekday phenomenon).

### Formal test: is the spatial pattern actually different?

Eyeballing two KDE panels side by side is suggestive but not a statistical test. `spatstat::segregation.test()` is built for exactly this: given a marked point pattern (here, the mark is the time category), it tests whether the *relative* spatial risk surface between the two marks is significantly non-constant — i.e. whether one category is genuinely more concentrated in some places than the other, beyond what you'd expect if every point's mark were just randomly relabelled.

```{r}
#| cache: true
seg_test <- function(sf_data, mark_col, sigma, nsim = 99, seed = 415) {
  co <- st_coordinates(sf_data)
  pm <- ppp(co[, 1], co[, 2], window = owin_bmr, marks = factor(sf_data[[mark_col]]))
  pm <- rjitter(pm, retry = TRUE, nsim = 1, radius = 5)
  set.seed(seed)
  segregation.test(pm, sigma = sigma, nsim = nsim)
}

seg_period  <- seg_test(acc_sf, "period",  sigma = 3000)
seg_daytype <- seg_test(acc_sf, "daytype", sigma = 3000)

seg_period
seg_daytype
```

I used a larger, fixed `sigma = 3000` m for this test specifically — `segregation.test()` calls `relrisk()` internally to estimate the spatial risk surface, and at the `bw.ppl()` scale (~750 m) that surface underflows to numerically unstable 0/1 values over large stretches of the study area (a `spatstat` warning flags this directly). A larger bandwidth is needed for a *relative-risk* surface to stay numerically well-behaved across an area this size than for a plain density surface — the two are estimating different things even though both use "sigma."

## Interpretation

Putting the continuous STKDE picture and the discrete formal test together, they agree and reinforce each other:

- The **hour-of-day STKDE** shows the hotspot's location and shape genuinely reshaping across the 24-hour cycle, not just dimming and brightening in place — and the discrete test confirms this is statistically real: **day vs. night, p = `r round(seg_period$p.value, 3)`**, significant spatial segregation. Where accidents happen genuinely shifts between day and night, not just how many.
- The **weekday/weekend discrete test gives the opposite answer — p = `r round(seg_daytype$p.value, 3)`**, not significant. Weekday and weekend accidents share essentially the same spatial footprint; only the volume changes (2,549 vs. 1,041). I didn't run a weekday/weekend STKDE for the same reason there's no continuous "weekday-ness" variable to feed it — day-of-week is inherently categorical, which is exactly the kind of split discrete testing is built for and STKDE isn't.
- The **month STKDE** adds a dimension neither of the day-of-week splits could show on its own: the hotspot doesn't relocate across the year, but it does intensify and sharpen from October through December, tracking the same months with the highest raw counts.

That maps onto a sensible real-world explanation: time-of-day changes *what kind* of driving is happening at a given location (peak-hour congestion at specific junctions vs. free-flowing night traffic), which is a spatial effect; weekday/weekend mostly changes *how much* traffic uses the same road network, a volume effect rather than a spatial one; and year-end changes both — more accidents, concentrated more tightly at the existing hotspot rather than spread out.

---

# Value-Added Analysis: Does Weather Affect Where Accidents Cluster?

The brief calls out `weather_condition`, `vehicle_type` and `presumed_cause` as optional extra angles. I went with **weather**, because it's the one with the cleanest natural comparison and enough data on both sides to trust the result: of the 3,590 points, 3,263 are logged as `clear` and 309 as `rainy` (the remaining `dark`/`foggy`/`other` categories are too small — 18 points combined — for a meaningful spatial test on their own). `presumed_cause` looked tempting too (alcohol-related accidents in particular), but the largest single non-"speeding"/"other" category, alcohol, is only 39 points BMR-wide in 2022 — too thin to say anything statistically defensible about its spatial pattern specifically.

```{r}
acc_sf |> st_drop_geometry() |> count(weather_condition, sort = TRUE)
```

## KDE by weather condition

Unlike the day/night comparison, I let each weather subset use its **own** `bw.ppl()` bandwidth here rather than sharing one — the two groups differ in size by an order of magnitude (3,263 vs. 309), and forcing the sparser rainy-day pattern through the same tight ~630 m bandwidth tuned for the dense clear-day pattern would just render as a scatter of unconnected dots, telling me nothing about where its own hotspots are.

```{r}
#| fig-cap: "KDE of accidents by weather condition, each panel on its own bandwidth and colour scale (see caveat below)."
clear_ppp <- make_ppp(acc_sf |> filter(weather_condition == "clear"))
rainy_ppp <- make_ppp(acc_sf |> filter(weather_condition == "rainy"))

sigma_clear <- bw.ppl(clear_ppp)
sigma_rainy <- bw.ppl(rainy_ppp)

kde_clear <- density(clear_ppp, sigma = sigma_clear)
kde_rainy <- density(rainy_ppp, sigma = sigma_rainy)

par(mfrow = c(1, 2))
plot(kde_clear, main = paste0("Clear (n=", clear_ppp$n, "), σ≈", round(sigma_clear), "m"))
plot(kde_rainy, main = paste0("Rainy (n=", rainy_ppp$n, "), σ≈", round(sigma_rainy), "m"))
```

Because the two panels use different bandwidths (~630 m vs. ~2,970 m) they aren't directly comparable by eye in the way the day/night pair was — the rainy panel is inherently smoother just from having fewer points to work with. What *is* comparable: clear-weather accidents trace the road network tightly, with the sharpest hotspot at the same central interchange seen throughout this document; rainy-weather accidents, even smoothed over a much wider radius, still concentrate most heavily in the east-central part of the study area rather than spreading evenly across it — a different part of the network than clear-weather's single dominant hotspot.

## Formal test

```{r}
#| cache: true
seg_weather <- seg_test(
  acc_sf |> filter(weather_condition %in% c("clear", "rainy")),
  "weather_condition", sigma = 3000
)
seg_weather
```

## Interpretation

`segregation.test()` returns **p = `r round(seg_weather$p.value, 3)`**, significant at the 5% level: clear-weather and rainy-weather accidents are not just different in volume (3,263 vs. 309, tracking the fact that it isn't raining most of the time) but occupy meaningfully different locations within the BMR. A plausible mechanism: rain doesn't just add slickness uniformly everywhere, it exposes drainage, visibility and road-surface weaknesses that are concentrated in particular stretches of the network — so the "risky when wet" locations aren't simply the same "risky all the time" locations scaled down. That's a genuinely different, actionable finding from the first-order KDE alone, which only ever showed one dominant hotspot regardless of conditions.

---

# Summary

## Key findings

1. **First-order**: 2022 BMR accidents are heavily concentrated along the expressway/arterial network, with the sharpest hotspot at a central-eastern interchange in Bangkok. Both fixed and adaptive KDE agree on the location.
2. **Second-order**: G, F, K and L function envelopes, plus a Clark-Evans R of 0.215 (p < 2.2 × 10⁻¹⁶), all confirm the clustering is statistically real and holds from the shortest to the longest spatial scales tested — this is not an artefact of random chance.
3. **Spatio-temporal**: STKDE (`sparr::spattemp.density()`, Chapter 6's method) shows the hotspot's location and shape reshaping continuously across the 24-hour cycle, and intensifying at the same location from October–December. The discrete formal test confirms day vs. night is a real spatial effect (`segregation.test` p ≈ `r round(seg_period$p.value,3)`) while weekday vs. weekend is not (p ≈ `r round(seg_daytype$p.value,3)`, volume-only).
4. **Value-added (weather)**: rainy-weather accidents cluster in a significantly different part of the BMR than clear-weather accidents do (p ≈ `r round(seg_weather$p.value,3)`), not just in smaller numbers.

## Limitations

- The BMR boundary is a name-based union of six provinces, not an official metro-area polygon — some edge effects near provincial borders are possible.
- 5% of BMR 2022 accident records have no coordinates and were dropped; if missingness isn't spatially random (e.g. concentrated in a particular reporting agency or district), the hotspots shown here could be mildly biased.
- 321 points shared exact duplicate coordinates before jittering, most likely reflecting reporting granularity (nearest landmark/km-marker) rather than true identical locations — jittering fixes the statistical degeneracy but the true precision of any individual point is uncertain.
- The day/night split (06:00–17:59 vs. 18:00–05:59) is a fixed clock-time boundary, not tied to actual daylight — a reasonable approximation at this latitude but not exact.

## Data sources

- Thailand Road Accident 2019–2022 dataset — [Kaggle](https://www.kaggle.com/datasets/thaweewatboy/thailand-road-accident-2019-2022)
- Thailand ADM1 province boundaries — [geoBoundaries](https://www.geoboundaries.org/)
